---
title: "KNN Classification Pipeline"
subtitle: "AMS 580 Team Project — Individual Contribution"
author: "Fei Qi"
date: "2026-04-26"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(message = FALSE, warning = FALSE)
```

```{r}
library(readr)
library(tidyverse)
library(caret)
library(class)
library(pROC)
```

```{r}
bank <- read.csv("../data/training_data.csv")
bank_1 <- read.csv("../data/testing_data.csv")
```

# Preprocessing
- duration, campaign, previous -> log scaling to account for skew
- duration should be removed from actual training data, can be used as a high-end baseline should we want to
- standardize the numerical data
- don't do anything yet with unknown. Can be a useful signal
- High multicollinearity with economic data, be mindful of it.
- "weak" singals from a few categorical variables to monitor
- target has strong class imbalance, use weighted models to counteract this

## Split the training and testing dataset
```{r}
train_index <- createDataPartition(bank$y, p = 0.75, list = FALSE)
training.samples <- bank$y %>% createDataPartition(p = 0.80, list = FALSE)
train  <- bank[training.samples, ]
test <- bank[-training.samples, ]
```

## Change pdays into categorical variable
```{r}
train$pdays_contacted_before <- ifelse(train$pdays == 999, "no", "yes")
test$pdays_contacted_before  <- ifelse(test$pdays == 999, "no", "yes")

train$pdays <- NULL
test$pdays  <- NULL
```

## Remove duration
Keeping a variable with duration should it be needed
```{r}
train$duration <- NULL
test$duration  <- NULL
```

## Convert categorical data to factors
```{r}
# Convert character columns in train to factors
train[] <- lapply(train, function(x) if(is.character(x)) as.factor(x) else x)

# Convert character columns in test to factors using train levels
for (col in names(train)) {
  if (is.factor(train[[col]])) {
    test[[col]] <- factor(test[[col]], levels = levels(train[[col]]))
  }
}
```

## Relevel for dummy variables created in logistic regression function
```{r}
train$job <- relevel(train$job, ref = "admin.")
train$marital <- relevel(train$marital, ref = "married")
train$education <- relevel(train$education, ref = "university.degree")
train$default <- relevel(train$default, ref = "no")
train$housing <- relevel(train$housing, ref = "no")
train$loan <- relevel(train$loan, ref = "no")
train$contact <- relevel(train$contact, ref = "cellular")
train$poutcome <- relevel(train$poutcome, ref = "nonexistent")
train$pdays_contacted_before <- relevel(train$pdays_contacted_before, ref = "no")

# Match test factor levels/order again
for (col in names(train)) {
  if (is.factor(train[[col]])) {
    test[[col]] <- factor(test[[col]], levels = levels(train[[col]]))
  }
}
```

## Apply log transform to relevant variables
```{r}
train$campaign <- log1p(train$campaign)
test$campaign  <- log1p(test$campaign)

train$previous <- log1p(train$previous)
test$previous  <- log1p(test$previous)
```

## Standardize the data
```{r}
num_cols <- sapply(train, is.numeric)
train_means <- sapply(train[, num_cols], mean)
train_sds   <- sapply(train[, num_cols], sd)

train_scaled <- train
train_scaled[, num_cols] <- scale(train[, num_cols],
                                 center = train_means,
                                 scale  = train_sds)

test_scaled <- test
test_scaled[, num_cols] <- scale(test[, num_cols],
                                center = train_means,
                                scale  = train_sds)
```

## KNN model
-KNN was tested as an additional non-parametric classification method to compare against logistic regression and tree-based models. 
- Unlike logistic regression, KNN does not assume a linear relationship between predictors and the outcome. 
-Instead, it classifies each observation based on nearby observations in the feature space
```{r}
# Convert categorical predictors into dummy variables
set.seed(123)
x_train_knn <- model.matrix(y ~ ., data = train_scaled)[, -1, drop = FALSE]
x_test_knn  <- model.matrix(y ~ ., data = test_scaled)[, -1, drop = FALSE]

x_train_knn <- as.data.frame(x_train_knn)
x_test_knn  <- as.data.frame(x_test_knn)

y_train_knn <- factor(train_scaled$y, levels = c("yes", "no"))
y_test_knn  <- factor(test_scaled$y, levels = c("yes", "no"))
```

# Creating balanced accuracy summary function

```{r}
# removing near-zero variables

nzv <- nearZeroVar(x_train_knn)

if (length(nzv) > 0) {
  x_train_knn <- x_train_knn[, -nzv]
  x_test_knn  <- x_test_knn[, -nzv]
}


set.seed(123)

train_knn_full <- data.frame(
  y = y_train_knn,
  x_train_knn
)

train_yes <- train_knn_full[train_knn_full$y == "yes", ]
train_no  <- train_knn_full[train_knn_full$y == "no", ]

n_yes <- nrow(train_yes)

train_no_sample <- train_no[sample(1:nrow(train_no), n_yes), ]

train_knn_balanced <- rbind(train_yes, train_no_sample)

# Shuffle rows
train_knn_balanced <- train_knn_balanced[sample(1:nrow(train_knn_balanced)), ]



```
# Train the knn model
```{r}
x_train_fast <- as.matrix(train_knn_balanced[, -1])
y_train_fast <- train_knn_balanced$y

x_test_fast <- as.matrix(x_test_knn)


k_values <- c(5, 11, 21, 31, 41)


knn_fast_results <- map_df(k_values, function(k_val) {
  
  pred <- knn(
    train = x_train_fast,
    test = x_test_fast,
    cl = y_train_fast,
    k = k_val,
    prob = TRUE
  )
  
  cm <- confusionMatrix(
    pred,
    y_test_knn,
    positive = "yes"
  )
  
  data.frame(
    k = k_val,
    accuracy = as.numeric(cm$overall["Accuracy"]),
    sensitivity = as.numeric(cm$byClass["Sensitivity"]),
    specificity = as.numeric(cm$byClass["Specificity"]),
    balanced_accuracy = as.numeric(cm$byClass["Balanced Accuracy"]),
    precision = as.numeric(cm$byClass["Precision"]),
    F1 = as.numeric(cm$byClass["F1"])
  )
})

knn_fast_results %>%
  arrange(desc(balanced_accuracy))



# choose best k

best_k <- knn_fast_results %>%
  arrange(desc(balanced_accuracy)) %>%
  slice(1) %>%
  pull(k)

best_k




# final kkn model

knn_final_pred <- knn(
  train = x_train_fast,
  test = x_test_fast,
  cl = y_train_fast,
  k = best_k,
  prob = TRUE
)

knn_final_cm <- confusionMatrix(
  knn_final_pred,
  y_test_knn,
  positive = "yes"
)

knn_final_cm


library(ggplot2)

p_k <- ggplot(knn_valid_results, aes(x = k, y = Bal_Acc)) +
  geom_line() +
  geom_point(size = 3) +
  geom_vline(xintercept = best_k, linetype = "dashed") +
  labs(
    title = "KNN Validation Performance by k",
    x = "Number of Neighbors (k)",
    y = "Balanced Accuracy"
  ) +
  theme_minimal()

p_k

ggsave(
  "../docs/images/knn_validation_by_k.png",
  plot = p_k,
  width = 8,
  height = 5,
  dpi = 300
)
```
#Final test metrics bar chart
```{r}
test_metrics <- knn_final_results %>%
  filter(Dataset == "Test") %>%
  select(Acc, Sens, Spec, Bal_Acc, F1, AUC) %>%
  pivot_longer(
    cols = everything(),
    names_to = "Metric",
    values_to = "Value"
  )

test_metrics$Metric <- factor(
  test_metrics$Metric,
  levels = c("Acc", "Sens", "Spec", "Bal_Acc", "F1", "AUC"),
  labels = c(
    "Accuracy",
    "Sensitivity",
    "Specificity",
    "Balanced Accuracy",
    "F1 Score",
    "AUC"
  )
)

p_metrics <- ggplot(test_metrics, aes(x = Metric, y = Value)) +
  geom_col() +
  geom_text(
    aes(label = sprintf("%.3f", Value)),
    vjust = -0.4
  ) +
  ylim(0, 1) +
  labs(
    title = "Final KNN Test Performance",
    x = NULL,
    y = "Score"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 35, hjust = 1)
  )

p_metrics

ggsave(
  "../docs/images/knn_test_metrics.png",
  plot = p_metrics,
  width = 9,
  height = 5,
  dpi = 300
)
```
# Test Data Result after finding the best k
```{r}
knn_test_pred <- knn(
  train = x_train_fast,
  test = x_test_fast,
  cl = y_train_fast,
  k = best_k,
  prob = TRUE
)

knn_test_pred <- factor(knn_test_pred, levels = c("yes", "no"))


#Test-set confusion matrix

knn_test_cm <- confusionMatrix(
  knn_test_pred,
  y_test_knn,
  positive = "yes"
)

knn_test_cm



#Test set table summary

knn_test_results <- data.frame(
  model = "KNN",
  k = best_k,
  accuracy = as.numeric(knn_test_cm$overall["Accuracy"]),
  sensitivity = as.numeric(knn_test_cm$byClass["Sensitivity"]),
  specificity = as.numeric(knn_test_cm$byClass["Specificity"]),
  precision = as.numeric(knn_test_cm$byClass["Precision"]),
  F1 = as.numeric(knn_test_cm$byClass["F1"]),
  balanced_accuracy = as.numeric(knn_test_cm$byClass["Balanced Accuracy"])
)

knn_test_results



# Confusion Matrix

library(yardstick)
library(ggplot2)
library(dplyr)

# Create dataframe in yardstick format
knn_df <- data.frame(
  truth = factor(y_test_knn, levels = c("no", "yes")),
  prediction = factor(knn_test_pred, levels = c("no", "yes"))
)







## for validation and test


train_index <- createDataPartition(bank$y, p = 0.75, list = FALSE)

train_model_raw <- bank[train_index, ]
valid_model_raw <- bank[-train_index, ]
test_model_raw  <- bank_1


# ------------------------------------------------------------
# 3. Preprocessing functions
# ------------------------------------------------------------

basic_clean <- function(df) {
  
  # Remove duration to avoid data leakage
  if ("duration" %in% names(df)) {
    df$duration <- NULL
  }
  
  # Convert pdays into binary contacted-before variable
  if ("pdays" %in% names(df)) {
    df$pdays_contacted_before <- ifelse(df$pdays == 999, "no", "yes")
    df$pdays_contacted_before <- factor(df$pdays_contacted_before)
    df$pdays <- NULL
  }
  
  # Log-transform skewed variables
  if ("campaign" %in% names(df)) {
    df$campaign <- log1p(df$campaign)
  }
  
  if ("previous" %in% names(df)) {
    df$previous <- log1p(df$previous)
  }
  
  # Convert character variables to factors
  df <- df %>%
    mutate(across(where(is.character), as.factor))
  
  # Make response variable consistent
  df$y <- factor(df$y, levels = c("no", "yes"))
  
  return(df)
}


# Clean raw sets
train_model_clean <- basic_clean(train_model_raw)
valid_model_clean <- basic_clean(valid_model_raw)
test_model_clean  <- basic_clean(test_model_raw)


# ------------------------------------------------------------
# 4. Match factor levels across train, validation, and test
# ------------------------------------------------------------

factor_cols <- names(train_model_clean)[sapply(train_model_clean, is.factor)]
factor_cols <- setdiff(factor_cols, "y")

for (col in factor_cols) {
  train_levels <- levels(train_model_clean[[col]])
  
  valid_model_clean[[col]] <- factor(valid_model_clean[[col]], levels = train_levels)
  test_model_clean[[col]]  <- factor(test_model_clean[[col]], levels = train_levels)
}


# ------------------------------------------------------------
# 5. Standardize numeric predictors using training parameters only
# ------------------------------------------------------------

numeric_cols <- names(train_model_clean)[sapply(train_model_clean, is.numeric)]

train_means <- sapply(train_model_clean[numeric_cols], mean, na.rm = TRUE)
train_sds   <- sapply(train_model_clean[numeric_cols], sd, na.rm = TRUE)

# Avoid division by zero
train_sds[train_sds == 0] <- 1

for (col in numeric_cols) {
  train_model_clean[[col]] <- (train_model_clean[[col]] - train_means[col]) / train_sds[col]
  valid_model_clean[[col]] <- (valid_model_clean[[col]] - train_means[col]) / train_sds[col]
  test_model_clean[[col]]  <- (test_model_clean[[col]]  - train_means[col]) / train_sds[col]
}


# ------------------------------------------------------------
# 6. Create dummy-variable matrices for KNN
# ------------------------------------------------------------

x_train_knn <- model.matrix(y ~ ., data = train_model_clean)[, -1, drop = FALSE]
x_valid_knn <- model.matrix(y ~ ., data = valid_model_clean)[, -1, drop = FALSE]
x_test_knn  <- model.matrix(y ~ ., data = test_model_clean)[, -1, drop = FALSE]

x_train_knn <- as.data.frame(x_train_knn)
x_valid_knn <- as.data.frame(x_valid_knn)
x_test_knn  <- as.data.frame(x_test_knn)

# Clean column names
colnames(x_train_knn) <- make.names(colnames(x_train_knn), unique = TRUE)
colnames(x_valid_knn) <- make.names(colnames(x_valid_knn), unique = TRUE)
colnames(x_test_knn)  <- make.names(colnames(x_test_knn), unique = TRUE)


# ------------------------------------------------------------
# 7. Align validation/test columns with training columns
# ------------------------------------------------------------

# Add missing columns to validation
missing_valid <- setdiff(colnames(x_train_knn), colnames(x_valid_knn))
for (col in missing_valid) {
  x_valid_knn[[col]] <- 0
}

# Add missing columns to test
missing_test <- setdiff(colnames(x_train_knn), colnames(x_test_knn))
for (col in missing_test) {
  x_test_knn[[col]] <- 0
}

# Remove extra columns and reorder
x_valid_knn <- x_valid_knn[, colnames(x_train_knn), drop = FALSE]
x_test_knn  <- x_test_knn[, colnames(x_train_knn), drop = FALSE]


# Response variables
y_train_knn <- factor(train_model_clean$y, levels = c("no", "yes"))
y_valid_knn <- factor(valid_model_clean$y, levels = c("no", "yes"))
y_test_knn  <- factor(test_model_clean$y, levels = c("no", "yes"))


# ------------------------------------------------------------
# 8. Remove near-zero variance predictors
# ------------------------------------------------------------

nzv <- nearZeroVar(x_train_knn)

if (length(nzv) > 0) {
  x_train_knn <- x_train_knn[, -nzv, drop = FALSE]
  x_valid_knn <- x_valid_knn[, -nzv, drop = FALSE]
  x_test_knn  <- x_test_knn[, -nzv, drop = FALSE]
}


# ------------------------------------------------------------
# 9. Create balanced training subset for KNN
# ------------------------------------------------------------

train_knn_full <- data.frame(
  y = y_train_knn,
  x_train_knn
)

train_yes <- train_knn_full[train_knn_full$y == "yes", ]
train_no  <- train_knn_full[train_knn_full$y == "no", ]

n_yes <- nrow(train_yes)

set.seed(123)
train_no_sample <- train_no[sample(1:nrow(train_no), n_yes), ]

train_knn_balanced <- rbind(train_yes, train_no_sample)
train_knn_balanced <- train_knn_balanced[sample(1:nrow(train_knn_balanced)), ]

x_train_fast <- as.matrix(train_knn_balanced[, -1])
y_train_fast <- factor(train_knn_balanced$y, levels = c("no", "yes"))

x_valid_fast <- as.matrix(x_valid_knn)
x_test_fast  <- as.matrix(x_test_knn)


# ------------------------------------------------------------
# 10. Helper function for KNN metrics
# ------------------------------------------------------------

evaluate_knn <- function(x_eval, y_eval, dataset_name, k_val) {
  
  pred_raw <- knn(
    train = x_train_fast,
    test = x_eval,
    cl = y_train_fast,
    k = k_val,
    prob = TRUE
  )
  
  # Extract KNN vote probability before factor conversion
  vote_prob <- attr(pred_raw, "prob")
  
  pred <- factor(as.character(pred_raw), levels = c("no", "yes"))
  y_eval <- factor(y_eval, levels = c("no", "yes"))
  
  # Convert vote probability to probability of "yes"
  yes_prob <- ifelse(
    pred == "yes",
    vote_prob,
    1 - vote_prob
  )
  
  cm <- confusionMatrix(
    pred,
    y_eval,
    positive = "yes"
  )
  
  auc_value <- pROC::auc(
    pROC::roc(
      response = y_eval,
      predictor = yes_prob,
      levels = c("no", "yes"),
      quiet = TRUE
    )
  )
  
  data.frame(
    Dataset = dataset_name,
    k = k_val,
    Bal_Acc = as.numeric(cm$byClass["Balanced Accuracy"]),
    AUC = as.numeric(auc_value),
    F1 = as.numeric(cm$byClass["F1"]),
    Sens = as.numeric(cm$byClass["Sensitivity"]),
    Spec = as.numeric(cm$byClass["Specificity"]),
    Acc = as.numeric(cm$overall["Accuracy"])
  )
}


# ------------------------------------------------------------
# 11. Choose best k using validation set only
# ------------------------------------------------------------

k_values <- c(5, 11, 21, 31, 41)

knn_valid_results <- map_df(k_values, function(k_val) {
  evaluate_knn(
    x_eval = x_valid_fast,
    y_eval = y_valid_knn,
    dataset_name = "Validation",
    k_val = k_val
  )
})

knn_valid_results %>%
  arrange(desc(Bal_Acc))

best_k <- knn_valid_results %>%
  arrange(desc(Bal_Acc)) %>%
  slice(1) %>%
  pull(k)

best_k


# ------------------------------------------------------------
# 12. Final validation and test results using best k
# ------------------------------------------------------------

knn_final_results <- bind_rows(
  evaluate_knn(x_valid_fast, y_valid_knn, "Validation", best_k),
  evaluate_knn(x_test_fast, y_test_knn, "Test", best_k)
)

knn_final_results



#Confusion

# Make sure this is the real held-out test set
test_model_raw <- bank_1

# Clean test set using the same function
test_model_clean <- basic_clean(test_model_raw)

# Match factor levels to training data
for (col in factor_cols) {
  train_levels <- levels(train_model_clean[[col]])
  test_model_clean[[col]] <- factor(test_model_clean[[col]], levels = train_levels)
}

# Standardize numeric columns using training means/sds
for (col in numeric_cols) {
  test_model_clean[[col]] <- (test_model_clean[[col]] - train_means[col]) / train_sds[col]
}

# Create test matrix
x_test_knn_real <- model.matrix(y ~ ., data = test_model_clean)[, -1, drop = FALSE]
x_test_knn_real <- as.data.frame(x_test_knn_real)

colnames(x_test_knn_real) <- make.names(colnames(x_test_knn_real), unique = TRUE)

# Add missing columns from training matrix
missing_test <- setdiff(colnames(x_train_knn), colnames(x_test_knn_real))
for (col in missing_test) {
  x_test_knn_real[[col]] <- 0
}

# Reorder to match training columns
x_test_knn_real <- x_test_knn_real[, colnames(x_train_knn), drop = FALSE]

# Remove the same near-zero variance columns
if (length(nzv) > 0) {
  x_test_knn_real <- x_test_knn_real[, -nzv, drop = FALSE]
}

# Final real test matrix and labels
x_test_fast_real <- as.matrix(x_test_knn_real)
y_test_knn_real <- factor(test_model_clean$y, levels = c("no", "yes"))

# Check size
nrow(x_test_fast_real)
length(y_test_knn_real)

knn_test_pred_raw <- knn(
  train = x_train_fast,
  test = x_test_fast,
  cl = y_train_fast,
  k = best_k,
  prob = TRUE
)

knn_test_pred <- factor(
  as.character(knn_test_pred_raw),
  levels = c("no", "yes")
)

# Final confusion matrix
knn_test_cm <- confusionMatrix(
  knn_test_pred,
  y_test_knn,
  positive = "yes"
)

knn_test_cm


```

#heatmap
```{r}
cm_table <- table(
  Actual = y_test_knn,
  Predicted = knn_test_pred
)

cm_df <- as.data.frame(cm_table)

p_cm <- ggplot(
  cm_df,
  aes(x = Predicted, y = Actual, fill = Freq)
) +
  geom_tile() +
  geom_text(
    aes(label = Freq),
    size = 6
  ) +
  labs(
    title = "KNN Test Confusion Matrix",
    x = "Predicted Class",
    y = "Actual Class"
  ) +
  theme_minimal()

p_cm

ggsave(
  "../docs/images/knn_confusion_matrix.png",
  plot = p_cm,
  width = 6,
  height = 5,
  dpi = 300
)
```