---
title: "KNN Classification Pipeline"
subtitle: "AMS 580 Team Project — Individual Contribution"
author: "Fei Qi"
date: "2026"
output:
  html_document:
    toc: true
    toc_float: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE
)
```

# Project Overview

This analysis presents my K-Nearest Neighbors (KNN) contribution to a five-person AMS 580 classification project.

The workflow uses the provided training dataset for model development, creates a separate validation subset for hyperparameter selection, and reserves the provided testing dataset for final evaluation.

Because the response variable is imbalanced, balanced accuracy is used as the primary validation metric. Accuracy, sensitivity, specificity, F1 score, and AUC are also reported.

# Libraries

```{r}
library(tidyverse)
library(caret)
library(class)
library(pROC)
library(readr)
```

# Load Data

```{r}
training_data <- read.csv("../data/training_data.csv")
testing_data  <- read.csv("../data/testing_data.csv")

dim(training_data)
dim(testing_data)
table(training_data$y)
```

# Train / Validation / Test Design

The provided training dataset is split into model-training and validation subsets. The provided testing dataset is kept separate and is not used to select `k`.

```{r}
set.seed(123)

train_index <- createDataPartition(
  training_data$y,
  p = 0.75,
  list = FALSE
)

train_raw <- training_data[train_index, ]
valid_raw <- training_data[-train_index, ]
test_raw  <- testing_data
```

# Preprocessing

The preprocessing workflow removes `duration`, converts `pdays` into a prior-contact indicator, applies log transformations to skewed count variables, converts character variables to factors, standardizes numeric predictors using training-set statistics only, and prepares dummy-variable matrices for KNN.

## Basic Cleaning

```{r}
basic_clean <- function(df) {

  # Remove duration to avoid using information unavailable before the outcome.
  if ("duration" %in% names(df)) {
    df$duration <- NULL
  }

  # Convert pdays into a binary prior-contact indicator.
  if ("pdays" %in% names(df)) {
    df$pdays_contacted_before <- ifelse(df$pdays == 999, "no", "yes")
    df$pdays_contacted_before <- factor(df$pdays_contacted_before)
    df$pdays <- NULL
  }

  # Log-transform skewed count variables.
  if ("campaign" %in% names(df)) {
    df$campaign <- log1p(df$campaign)
  }

  if ("previous" %in% names(df)) {
    df$previous <- log1p(df$previous)
  }

  # Convert character variables to factors.
  df <- df %>%
    mutate(across(where(is.character), as.factor))

  # Standardize response levels.
  df$y <- factor(df$y, levels = c("no", "yes"))

  df
}

train_clean <- basic_clean(train_raw)
valid_clean <- basic_clean(valid_raw)
test_clean  <- basic_clean(test_raw)
```

## Match Factor Levels

Validation and test categorical variables are aligned to the factor levels learned from the training subset.

```{r}
factor_cols <- names(train_clean)[sapply(train_clean, is.factor)]
factor_cols <- setdiff(factor_cols, "y")

for (col in factor_cols) {
  train_levels <- levels(train_clean[[col]])

  valid_clean[[col]] <- factor(
    valid_clean[[col]],
    levels = train_levels
  )

  test_clean[[col]] <- factor(
    test_clean[[col]],
    levels = train_levels
  )
}
```

## Standardize Numeric Predictors

Numeric means and standard deviations are estimated from the training subset only and then applied to validation and test data.

```{r}
numeric_cols <- names(train_clean)[sapply(train_clean, is.numeric)]

train_means <- sapply(
  train_clean[numeric_cols],
  mean,
  na.rm = TRUE
)

train_sds <- sapply(
  train_clean[numeric_cols],
  sd,
  na.rm = TRUE
)

# Avoid division by zero for constant predictors.
train_sds[train_sds == 0] <- 1

for (col in numeric_cols) {
  train_clean[[col]] <- (
    train_clean[[col]] - train_means[col]
  ) / train_sds[col]

  valid_clean[[col]] <- (
    valid_clean[[col]] - train_means[col]
  ) / train_sds[col]

  test_clean[[col]] <- (
    test_clean[[col]] - train_means[col]
  ) / train_sds[col]
}
```

# Feature Preparation for KNN

## Dummy Encoding

```{r}
x_train <- model.matrix(
  y ~ .,
  data = train_clean
)[, -1, drop = FALSE]

x_valid <- model.matrix(
  y ~ .,
  data = valid_clean
)[, -1, drop = FALSE]

x_test <- model.matrix(
  y ~ .,
  data = test_clean
)[, -1, drop = FALSE]

x_train <- as.data.frame(x_train)
x_valid <- as.data.frame(x_valid)
x_test  <- as.data.frame(x_test)

colnames(x_train) <- make.names(
  colnames(x_train),
  unique = TRUE
)

colnames(x_valid) <- make.names(
  colnames(x_valid),
  unique = TRUE
)

colnames(x_test) <- make.names(
  colnames(x_test),
  unique = TRUE
)
```

## Align Predictor Columns

```{r}
missing_valid <- setdiff(
  colnames(x_train),
  colnames(x_valid)
)

for (col in missing_valid) {
  x_valid[[col]] <- 0
}

missing_test <- setdiff(
  colnames(x_train),
  colnames(x_test)
)

for (col in missing_test) {
  x_test[[col]] <- 0
}

x_valid <- x_valid[
  ,
  colnames(x_train),
  drop = FALSE
]

x_test <- x_test[
  ,
  colnames(x_train),
  drop = FALSE
]

y_train <- factor(
  train_clean$y,
  levels = c("no", "yes")
)

y_valid <- factor(
  valid_clean$y,
  levels = c("no", "yes")
)

y_test <- factor(
  test_clean$y,
  levels = c("no", "yes")
)
```

## Remove Near-Zero-Variance Predictors

```{r}
nzv <- nearZeroVar(x_train)

if (length(nzv) > 0) {
  x_train <- x_train[
    ,
    -nzv,
    drop = FALSE
  ]

  x_valid <- x_valid[
    ,
    -nzv,
    drop = FALSE
  ]

  x_test <- x_test[
    ,
    -nzv,
    drop = FALSE
  ]
}
```

# Class Imbalance Handling

The majority class is undersampled in the model-training subset so that KNN is trained on a balanced sample.

```{r}
train_knn_full <- data.frame(
  y = y_train,
  x_train
)

train_yes <- train_knn_full[
  train_knn_full$y == "yes",
]

train_no <- train_knn_full[
  train_knn_full$y == "no",
]

n_yes <- nrow(train_yes)

set.seed(123)

train_no_sample <- train_no[
  sample(
    seq_len(nrow(train_no)),
    n_yes
  ),
]

train_knn_balanced <- rbind(
  train_yes,
  train_no_sample
)

train_knn_balanced <- train_knn_balanced[
  sample(
    seq_len(nrow(train_knn_balanced))
  ),
]

x_train_knn <- as.matrix(
  train_knn_balanced[, -1]
)

y_train_knn <- factor(
  train_knn_balanced$y,
  levels = c("no", "yes")
)

x_valid_knn <- as.matrix(x_valid)
x_test_knn  <- as.matrix(x_test)

table(y_train_knn)
```

# KNN Evaluation Function

```{r}
evaluate_knn <- function(
  x_eval,
  y_eval,
  dataset_name,
  k_val
) {

  pred_raw <- knn(
    train = x_train_knn,
    test = x_eval,
    cl = y_train_knn,
    k = k_val,
    prob = TRUE
  )

  vote_prob <- attr(
    pred_raw,
    "prob"
  )

  pred <- factor(
    as.character(pred_raw),
    levels = c("no", "yes")
  )

  y_eval <- factor(
    y_eval,
    levels = c("no", "yes")
  )

  # class::knn returns the vote proportion for the predicted class.
  # Convert it into a score for the positive ("yes") class.
  yes_prob <- ifelse(
    pred == "yes",
    vote_prob,
    1 - vote_prob
  )

  cm <- confusionMatrix(
    pred,
    y_eval,
    positive = "yes"
  )

  auc_value <- pROC::auc(
    pROC::roc(
      response = y_eval,
      predictor = yes_prob,
      levels = c("no", "yes"),
      quiet = TRUE
    )
  )

  data.frame(
    Dataset = dataset_name,
    k = k_val,
    Bal_Acc = as.numeric(
      cm$byClass["Balanced Accuracy"]
    ),
    AUC = as.numeric(auc_value),
    F1 = as.numeric(
      cm$byClass["F1"]
    ),
    Sens = as.numeric(
      cm$byClass["Sensitivity"]
    ),
    Spec = as.numeric(
      cm$byClass["Specificity"]
    ),
    Acc = as.numeric(
      cm$overall["Accuracy"]
    )
  )
}
```

# Model Selection

Multiple values of `k` are evaluated on the validation subset only. The value with the highest balanced accuracy is selected for the final model.

```{r}
k_values <- c(
  5,
  11,
  21,
  31,
  41
)

knn_valid_results <- map_df(
  k_values,
  function(k_val) {
    evaluate_knn(
      x_eval = x_valid_knn,
      y_eval = y_valid,
      dataset_name = "Validation",
      k_val = k_val
    )
  }
)

knn_valid_results <- knn_valid_results %>%
  arrange(desc(Bal_Acc))

knn_valid_results

best_k <- knn_valid_results %>%
  slice(1) %>%
  pull(k)

best_k
```

# Final Held-Out Test Evaluation

The selected value of `k` is evaluated once on the provided held-out test dataset.

```{r}
knn_final_results <- bind_rows(
  evaluate_knn(
    x_valid_knn,
    y_valid,
    "Validation",
    best_k
  ),
  evaluate_knn(
    x_test_knn,
    y_test,
    "Test",
    best_k
  )
)

knn_final_results
```

## Final Test Predictions

```{r}
knn_test_pred_raw <- knn(
  train = x_train_knn,
  test = x_test_knn,
  cl = y_train_knn,
  k = best_k,
  prob = TRUE
)

knn_test_pred <- factor(
  as.character(knn_test_pred_raw),
  levels = c("no", "yes")
)

knn_test_cm <- confusionMatrix(
  knn_test_pred,
  y_test,
  positive = "yes"
)

knn_test_cm
```

# Visualizations

Create the output directory if it does not already exist.

```{r}
dir.create(
  "../docs/images",
  recursive = TRUE,
  showWarnings = FALSE
)
```

## Validation Performance by k

```{r}
p_k <- ggplot(
  knn_valid_results,
  aes(
    x = k,
    y = Bal_Acc
  )
) +
  geom_line() +
  geom_point(size = 3) +
  geom_vline(
    xintercept = best_k,
    linetype = "dashed"
  ) +
  annotate(
    "text",
    x = best_k,
    y = max(knn_valid_results$Bal_Acc),
    label = paste0(
      "Selected k = ",
      best_k
    ),
    vjust = -1,
    hjust = 0.5
  ) +
  labs(
    title = "KNN Validation Performance by k",
    x = "Number of Neighbors (k)",
    y = "Balanced Accuracy"
  ) +
  theme_minimal()

p_k

ggsave(
  "../docs/images/knn_validation_by_k.png",
  plot = p_k,
  width = 8,
  height = 5,
  dpi = 300
)
```

## Final Test Metrics

```{r}
test_metrics <- knn_final_results %>%
  filter(Dataset == "Test") %>%
  select(
    Acc,
    Sens,
    Spec,
    Bal_Acc,
    F1,
    AUC
  ) %>%
  pivot_longer(
    cols = everything(),
    names_to = "Metric",
    values_to = "Value"
  )

test_metrics$Metric <- factor(
  test_metrics$Metric,
  levels = c(
    "Acc",
    "Sens",
    "Spec",
    "Bal_Acc",
    "F1",
    "AUC"
  ),
  labels = c(
    "Accuracy",
    "Sensitivity",
    "Specificity",
    "Balanced Accuracy",
    "F1 Score",
    "AUC"
  )
)

p_metrics <- ggplot(
  test_metrics,
  aes(
    x = Metric,
    y = Value
  )
) +
  geom_col() +
  geom_text(
    aes(
      label = sprintf(
        "%.3f",
        Value
      )
    ),
    vjust = -0.4
  ) +
  coord_cartesian(
    ylim = c(0, 1)
  ) +
  labs(
    title = paste0(
      "Final KNN Test Performance (k = ",
      best_k,
      ")"
    ),
    x = NULL,
    y = "Score"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(
      angle = 35,
      hjust = 1
    )
  )

p_metrics

ggsave(
  "../docs/images/knn_test_metrics.png",
  plot = p_metrics,
  width = 9,
  height = 5,
  dpi = 300
)
```

## Confusion Matrix Heatmap

```{r}
cm_table <- table(
  Actual = y_test,
  Predicted = knn_test_pred
)

cm_df <- as.data.frame(
  cm_table
)

p_cm <- ggplot(
  cm_df,
  aes(
    x = Predicted,
    y = Actual,
    fill = Freq
  )
) +
  geom_tile() +
  geom_text(
    aes(label = Freq),
    size = 6
  ) +
  guides(fill = "none") +
  labs(
    title = paste0(
      "Held-Out Test Confusion Matrix (k = ",
      best_k,
      ")"
    ),
    x = "Predicted Class",
    y = "Actual Class"
  ) +
  theme_minimal()

p_cm

ggsave(
  "../docs/images/knn_confusion_matrix.png",
  plot = p_cm,
  width = 6,
  height = 5,
  dpi = 300
)
```

# Conclusion

This pipeline keeps model selection and final evaluation separate: `k` is selected using validation balanced accuracy, and the selected KNN model is then evaluated on the held-out test dataset.

The workflow also standardizes numeric predictors using training-set statistics only, aligns encoded predictors across datasets, removes near-zero-variance features, and balances the model-training subset before fitting KNN.
